Practical Effects of Integrating Temperature with Strang Split Reactions

### 2021-02-23 strang/paper.tex
fix typo
new notebook

When we are reacting, we need to evolve the temperature, which we can write as:
equation:
\begin{equation}
\rho c_v \frac{DT}{Dt} = \rho \left (\frac{p}{\rho^2} - e_\rho \right ) \frac{D\rho}{Dt} + \rho \Sdot
\end{equation}
(this form neglects the change in temperature due to composition
changes from reactions, see \citealt{ABNZ:III}, which assumes chemical
equilibrium).

@@

During reactions, we neglect the hydrodynamics terms,
so the reactive system updates according to:
\begin{align}
\frac{D\rho}{Dt} &= 0 \\
\frac{DX_k}{Dt} &= \omegadot_k \label{eq:species} \\
\frac{DT}{Dt} &= \frac{\Sdot}{c_v}
\end{align}

There are several different approximations that people make in the
literature to this operator-split reacting system, in order to make
integrating the reaction system computationally less expensive:
\begin{itemize}
\item Neglect temerature evolution completely in the burn, only
  evolving Eq.~\ref{eq:species}.  This is the method used in
  \citet{flash}. \MarginPar{other codes?}
\item Evolve the temperature equation, but ``freeze'' the specific
  heat, $c_v$, at the start of integration to avoid potentially
  expensive equation of state calls.  This was discussed in
  \cite{Bell:2004} and is the default method in \castro.
\item Evolve the full system, calling the equation of state each time
  we evaluate the righthand side of the system to get an updated
  $c_v$.  This is an option in \castro.
\end{itemize}
Depending on how vigorous the burning is and how much the temperature
changes during a hydrodynamic timestep, one or more of these
approximations may be reasonable.  For explosive reactions, we expect
that evolving the full system will be required.  The goal of this
note is to try to quantify the convergence of a reacting hydrodynamics
code with these different approximations.

>>>>

When we are reacting, we can look at internal energy:
\begin{equation}
\rho \frac{De}{Dt} + p \nabla \cdot \Ub = \rho \Sdot
\end{equation}
or alternately, we can evolve the temperature, which we can write as:
\begin{equation}
\rho c_v \frac{DT}{Dt} = \rho \left (\frac{p}{\rho^2} - e_\rho \right ) \frac{D\rho}{Dt} + \rho \Sdot
\end{equation}
(this form neglects the change in temperature due to composition
changes from reactions, see \citealt{ABNZ:III}, which assumes chemical
equilibrium).

@@

During reactions, we neglect the hydrodynamics terms,
so the reactive system updates according to:
\begin{align}
\frac{D\rho}{Dt} &= 0 \\
\frac{DX_k}{Dt} &= \omegadot_k \label{eq:species}
\end{align}
with either
\begin{equation}
\frac{DT}{Dt} = \frac{\Sdot}{c_v}
\end{equation}
or
\begin{equation}
\frac{De}{Dt} = \Sdot
\end{equation}

There are several different approximations that people make in the
literature to this operator-split reacting system, in order to make
integrating the reaction system computationally less expensive:
\begin{itemize}
\item Evolve $(X_k)$ only.  This neglects temerature evolution
  completely in the burn, only evolving Eq.~\ref{eq:species}.  This is
  the method used in \citet{flash}. \MarginPar{other codes?}

\item Evolve $(X_k, T)$, but ``freeze'' the specific
  heat, $c_v$, at the start of integration to avoid potentially
  expensive equation of state calls.  This was discussed in
  \cite{Bell:2004} and is the default method in \castro.

\item Evolve $(X_k, T)$, calling the equation of state each time
  we evaluate the righthand side of the system to get an updated
  $c_v$.  This is an option in \castro.

\item Evolve $(X_k, e)$, and get $T$ from $e$ using the equation of
  state.
\end{itemize}
Depending on how vigorous the burning is and how much the temperature
changes during a hydrodynamic timestep, one or more of these
approximations may be reasonable.  For explosive reactions, we expect
that evolving the full system will be required.  The goal of this
note is to try to quantify the convergence of a reacting hydrodynamics
code with these different approximations.
